# Propensity score methods {#iptw-application}

## Background

The **Propensity Score (PS)** is utilized to emulate the randomization process, with the goal of balancing the population's characteristics across the groups of interest. Unlike multivariable regression, which focuses on estimating the effect at the individual level (*subject-specific effect*), the PS method estimates the effect at the population level (*population-average effect*). In the context of models with a non-linear link (typically when Odds Ratio [OR] or Hazard Ratio [HR] are reported), these two effects differ [@austin_conditioning_2007].

Statistically, the PS is defined as the conditional probability of receiving the exposure of interest based on observed baseline characteristics. It is typically estimated using multivariable logistic regression, where the exposure of interest serves as the dependent variable and the baseline characteristics act as independent variables. Once calculated, it can be employed in various approaches, including matching, stratification, **Inverse Probability of Treatment Weighting** (IPTW), and covariate adjustment [@austin_introduction_2011].

The scenario 1 discussed in Chapter \@ref(motivating-example) will be applied to illustrate the application of IPTW.

## Identify confounders

The baseline characteristics incorporated in the logistic regression must be defined as **confounders - variables that influence both treatment assignment and the outcome without being part on the causal pathway**. Other types of variables, such as **mediators or colliders**, may also be present; however, adjusting for these can introduce bias [@mackinnon_unification_2021].

**Directed Acyclic Graphs** (DAGs) are usually recommended to depict causal relationships, ideally, clinical experts should be involved in their development as they more likely to know involved clinical pathways, interacting with the disease of interest [@hernan_causal_nodate; @tennant_use_2021]. DAGs can be created using the DAGitty [website](https://www.dagitty.net/) or dagitty R [package](https://github.com/jtextor/dagitty).

In our example, we identified age at diagnosis, gender, Beck score and socioeconomic status as potential confounders of the association between treatment exposure and relapse. The associated DAG is represented in Appendix \@ref(dag).

## IPTW implementation

This code implements the IPTW approach using the **stabilized Average Treatment Effect (ATE)** estimand [@austin_moving_2015], detailed in the following equation:

$$
IPTW_{Stabilized, ATE} = \frac{X*Pr(X=1)}{e}+\frac{(1-X)*Pr(X=0)}{1-e} 
$$

Here, the PS is defined as $e = \Pr ({X}=1|{Z}={z})$ and $X$ and $Z$ are the treatment status (0 = Sertralex and 1 = Duloxyn) and observed confunders (gender, age, SES level, and Beck score), respectively.

This models the treatment effect across the entire population likely to receive either Sertralex or Duloxyn. Additional details on causal estimands (ATE/ATT) are provided in Appendix \@ref(causal).

```{r iptw, message=FALSE}
#Perform weighting
W <- WeightIt::weightit(TREAT ~ AGE + GENDER + SOCIO_ECO + BECK,
                            data = data,
                            method = "glm",
                            estimand = "ATE",
                            stabilize = TRUE)
data$weights <- W$weights
data$ps <- W$ps
```

## Balance Diagnostics

To evaluate balance after weighting, we applied the typical recommendation by using the Standardized Mean Difference (SMD) and checking whether the difference in SMD is less than 0.1 [@austin_balance_2009; @normand_validating_2001]. A difference below this threshold indicates that the characteristics of interest are balanced between the groups.

```{r love_plot}
love.plot(W, 
          binary = "std", 
          thresholds = c(m = .1),
          sample.names = c("Before weighting", "After weighting"))
```

To summarize, a table before and after weighting is also represented according to treatment status.

```{r table_weighting, warning=FALSE}
#Before weighting
t1 <- data %>%
  select(AGE, GENDER, SOCIO_ECO, BECK, TREAT) %>%
  tbl_summary(by = TREAT, 
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              label = list(AGE ~ "Age (at diagnosis)",
                           GENDER ~ "Gender",
                           SOCIO_ECO ~ "SES level (1-5)",
                           BECK ~ "Beck score (0-63)"),
              digits = all_continuous() ~ 1) %>%
  add_difference(test = everything() ~ "smd") %>%
  modify_header(label ~ "**Variable**", 
                estimate ~ "**SMD**", 
                stat_1 ~ paste0("**Sertralex (N=",nrow(data[which(data$TREAT=="Sertralex"),]),")**"), 
                stat_2 ~ paste0("**Duloxyn (N=",nrow(data[which(data$TREAT=="Duloxyn"),]),")**"))
data_selected <- data %>%
  select(AGE, GENDER, SOCIO_ECO, BECK, TREAT)

#After weighting
data_svy <- svydesign(ids = ~1, data = data_selected, weights = ~W$weights)
t2 <- data_svy %>%
  tbl_svysummary(by = TREAT, 
                 statistic = list(all_continuous() ~ "{mean} ({sd})"),
                 label = list(AGE ~ "Age (at diagnosis)",
                           GENDER ~ "Gender",
                           SOCIO_ECO ~ "SES level (1-5)",
                           BECK ~ "Beck score (0-63)"),
                 digits = all_continuous() ~ 1) %>%
  add_difference(test = everything() ~ "smd") %>%
  modify_header(label ~ "**Variable**", 
                estimate ~ "**SMD**", 
                stat_1 ~ paste0("**Sertralex (N=",round(ESS(W$weights[W$treat=="Sertralex"]),0),")**"), 
                stat_2 ~ paste0("**Duloxyn (N=",round(ESS(W$weights[W$treat=="Duloxyn"]),0),")**")) 

for (int_col in c("modify_stat_N", "modify_stat_n")) {
  t1$table_styling$header[[int_col]] <-
    t1$table_styling$header[[int_col]] |> as.numeric()
}

tbl_merge(tbls = list(t1, t2),
          tab_spanner = c(
            paste0("**Raw (N=",nrow(data),")**"), 
            paste0("**Weighted (ESS=",round(ESS(W$weights),0),")**")
            )
          )%>%
  modify_table_body(~ .x %>% select(-conf.low_1, -conf.low_2))
```

After weighting, observed confunders were above the recommended threshold. If, after weighting, the observed confounders exceed the recommended threshold, the model can be respecified (e.g. making a continuous variable categorical or dichotomous, including higher order terms or splines of variables, ...). Otherwise, variables with remaining imbalance should ideally be adjusted for when modeling the final outcome, using a doubly robust approach (i.e., adjusting for confounders in both the treatment and outcome models) [@funk_doubly_2011; @garrido_methods_2014].

Next, to ensure that the positivity assumption is met, we plotted the propensity score distributions to assess the degree of overlap between individuals receiving Sertralex and Duloxyn. Our goal is to establish sufficient common support, meaning all subjects have a non-zero probability of receiving either treatment.

```{r ps}
#Display PS
plot1<-ggplot2::ggplot(data, mapping = aes(x = ps))+
  ggplot2::geom_density(colour="red")+
  ggplot2::geom_density(mapping=aes(group = TREAT, fill = TREAT),alpha=0.4)+
  ggtitle("Raw")+
  scale_fill_manual(name="Treatment",labels=c("Sertralex","Duloxyn"),values=c("#FFDB6D","#00AFBB"))+
  xlab("Propensity Score (PS)")+
  ylab("Density")+
  xlim(0,1)+
  labs(caption="")+
  theme_minimal()
  
#After weighting
plot2<-ggplot2::ggplot(data, mapping = aes(x = ps, weight = weights))+
  ggplot2::geom_density(colour="red")+
  ggplot2::geom_density(mapping=aes(group = TREAT, fill = TREAT),alpha=0.4)+
  ggtitle("Weighted")+
  scale_fill_manual(name="Treatment",labels=c("Sertralex","Duloxyn"),values=c("#FFDB6D","#00AFBB"))+
  xlab("Propensity Score (PS)")+
  ylab("Density")+
  xlim(0,1)+
  labs(caption="*the red line indicates the PS distribution for the overall population")+
  theme_minimal()
  
ggarrange(plot1,plot2,common.legend = T)
```

We observed that after weighting the two distributions curves align, indicating the dataset is balanced on average. 

## Outcome analysis

This section details the outcome analyis. First, unadjusted and weighted Kaplan-Meier curves [@cole_adjusted_2004], stratified on treatment group, are compared.

```{r km, warning=FALSE}
#Implement unadjusted model
unadjusted<- survfit(Surv(TIME_TO_EVENT, EVENT)~TREAT, data = data)
#Implement IPTW adjusted model
weighted<- survfit(Surv(TIME_TO_EVENT, EVENT)~TREAT, data = data, weights = weights)
#Combine both
fit<-list(unadjusted,weighted)

# Function to format survival probabilities at a given time point
format_median_time <- function(surv_object, treatment_index) {
  
  # Extract the median survival time for the given treatment index
  median_surv <- sprintf("%.2f", round(summary(surv_object)$table[treatment_index, "median"], 2))
  lower_ci <- sprintf("%.2f", round(summary(surv_object)$table[treatment_index, "0.95LCL"], 2))
  upper_ci <- sprintf("%.2f", round(summary(surv_object)$table[treatment_index, "0.95UCL"], 2))
  
  return(paste0(median_surv, " [", lower_ci, "; ", upper_ci, "]"))
  
}

# Usage for raw and IPTW median time
unadjusted_treatment_0 <- format_median_time(unadjusted, 1)
unadjusted_treatment_1 <- format_median_time(unadjusted, 2)
weighted_treatment_0 <- format_median_time(weighted, 1)
weighted_treatment_1 <- format_median_time(weighted, 2)

ggsurvplot_combine(fit, 
                   data = data,
                   legend.title = "",
                   legend.labs = c("Unadjusted: Sertralex", 
                                   "Unadjusted: Duloxyn",
                                   "Weighted: Sertralex", 
                                   "Weighted: Duloxyn"),
                   palette = c("#ff331a","#ff80a6","#FFDB6D","#27d4ad"),
                   conf.int = TRUE,
                   xlab = "Time (in days)",
                   ggtheme = theme_bw())
```

As outlined, the median time to relapse differs depending on the analysis and treatment group considered. In the unadjusted analysis, we observed a median survival time of `r unadjusted_treatment_0` days for Sertralex, compared to `r unadjusted_treatment_1` days for Duloxyn. In the weighted analysis (where weights are used to create a pseudo population), these differences were reduced, median survival time of `r weighted_treatment_0` days for Sertralex and `r weighted_treatment_1` days for Duloxyn.

Second, unadjusted, multivariable, and weighted Cox are displayed.

```{r cox}
# Fit the Cox models
unadjusted <- coxph(Surv(TIME_TO_EVENT, EVENT) ~ TREAT, data = data)
multivariable <- coxph(Surv(TIME_TO_EVENT, EVENT) ~ TREAT + AGE + GENDER + SOCIO_ECO + BECK, data = data)
weighted <- coxph(Surv(TIME_TO_EVENT, EVENT) ~ TREAT, weights = weights, data = data)

# Create regression tables
unadjusted_table <- unadjusted %>% 
  tbl_regression(exponentiate = TRUE,
                 label = list(TREAT ~ "Treatment (unadjusted)"))

multivariable_table <- multivariable %>% 
  tbl_regression(exponentiate = TRUE,
                 label = list(TREAT ~ "Treatment (multivariable regression)"))%>%
  modify_table_body(
    ~ .x %>% dplyr::filter(variable == "TREAT")
  )

weighted_table <- weighted %>% 
  tbl_regression(exponentiate = TRUE,
                 label = list(TREAT ~ "Treatment (weighted)"))%>%
  modify_table_body(
    ~ .x %>% dplyr::filter(variable == "TREAT")
  )

# Function to extract HR and CI for a specific variable
extract_hr_ci <- function(model) {
  conf <- confint(model)
  hr <- sprintf("%.2f", exp(model$coefficients)[1])
  ci_lower <- sprintf("%.2f", exp(conf)[1,1])
  ci_upper <- sprintf("%.2f", exp(conf)[1,2])
  return(paste0("HR = ", hr, " [", ci_lower, "; ", ci_upper, "]"))
}

# Extract and format HR and CI for TREAT
unadjusted_hr_ci <- extract_hr_ci(unadjusted)
multivariable_hr_ci <- extract_hr_ci(multivariable)
weighted_hr_ci <- extract_hr_ci(weighted)

# Stack the regression tables
tbl_stack(list(unadjusted_table, multivariable_table, weighted_table))
```

The results indicate that the treatment effect varies depending on the analysis and treatment group considered. In the unadjusted analysis, the hazard ratio (`r unadjusted_hr_ci`) suggests that Duloxyn significantly increases the risk of relapse. In contrast, the multivariable model reveals the opposite findings with a hazard ratio of `r multivariable_hr_ci`. Meanwhile, the IPTW analysis yielded non-significant results with `r weighted_hr_ci`. Finally, the slight differences between the multivariable and IPTW models demonstrate that conditional and marginal effects are not strictly equivalent.
