# Clone-Censor-Weight method {#clone-censor-weight}

## Background

Another challenge common in RWD studies, in addition to confounding, is immortal time bias. This bias often occurs when there is a time difference between the beginning of follow-up (time zero) and the initiation of treatment, this is especially true when this systematically differs across groups.

For instance, if Sertralex patients begin treatment later than Duloxyn individuals, a form of immortal time occurs. In other words, individuals on Duloxyn might experience an event sooner, simply due to the earlier start of their follow-up period. This difference in initiation timing may not be random and could, in fact, be influenced by various factors that need to be considered into the analysis.

To address both confounding and immortal time bias, the Clone-Censor-Weight (CCW) approach was introduced [@hernan_specifying_2016; @maringe_reflection_2020]. Three common scenarios were distinguished for the use of CCW: grace periods-a period during which treatment initiation can happen, static time-related strategies, and dynamic strategies [@zhao_versatility_2021].

The scenario 2 discussed in Chapter \@ref(motivating-example) corresponds to the grace period scenario and will be applied to illustrate the application of CCW .

## CCW implementation

This code implements the CCW approach.

### Cloning

The cloning approach involves creating a duplicate of the initial dataset, where each clone is assigned the opposite treatment.

```{r clone, message=FALSE}
# Step 1 cloning
# Make a copy of original data
original_data <- data
original_data <- original_data %>%
                  mutate(SOCIO_ECO_2 = ifelse(SOCIO_ECO == "Low", 1, 0),
                         SOCIO_ECO_3 = ifelse(SOCIO_ECO == "Moderate", 1, 0),
                         SOCIO_ECO_4 = ifelse(SOCIO_ECO == "High", 1, 0),
                         SOCIO_ECO_5 = ifelse(SOCIO_ECO == "Very high", 1, 0)) %>%
  select(-SOCIO_ECO)

# Create cloned data
cloned_data <- original_data %>%
  mutate(
    ID = ID + 1000,                   # Modify ID to differentiate cloned records
    TREAT = ifelse(TREAT == "Duloxyn", "Sertralex", "Duloxyn")
  )         # Remove TIME_TO_TREAT for clones
```

### Censoring

Next, the censoring pattern must be established in the dataset. In the original dataset, no censoring is applied, whereas in the cloned dataset, censoring is set at the time of treatment initiation. The two datasets are then merged and reformatted to include risk intervals for each individual, defined with start and stop times.

```{r censor, message=FALSE}
# Step 2 censoring
original_data <- original_data %>%
  mutate(
    CENSOR = 0,                       # Indicator for uncensored data in the original dataset
    TIME_TO_CENSOR = TIME_TO_TREAT    # Initial censoring time is set to treatment time
  ) %>%
  select(-TIME_TO_TREAT)               # Remove TIME_TO_TREAT for simplicity

cloned_data <- cloned_data %>%
  mutate(
    EVENT = 0,                        # No relapse events for cloned records
    CENSOR = 1,                       # Cloned records are censored
    TIME_TO_CENSOR = TIME_TO_TREAT    # Use TIME_TO_EVENT as censoring time for clones
  ) %>%
  select(-TIME_TO_TREAT)               # Remove TIME_TO_TREAT for clones

# Combine original and cloned datasets
combined_data <- bind_rows(original_data, cloned_data) %>%
  mutate(
    TIME_TO_EVENT = TIME_TO_EVENT + 0.001,  # Add small increments to avoid duplicate times
    TIME_TO_CENSOR = TIME_TO_CENSOR + 0.001, # Add small increments for censoring times
    TSTART = 0                              # Set start time for all records
  )

# Identify unique times for events and censoring
t_events <- sort(unique(combined_data$TIME_TO_EVENT))  # Unique event times
t_censor <- sort(unique(combined_data$TIME_TO_CENSOR)) # Unique censoring times

# Reshape data for events using survSplit
combined_data_events <- survSplit(
  Surv(TSTART, TIME_TO_EVENT, EVENT) ~ ., 
  data = combined_data, 
  cut = t_events, 
  id = "id"
) %>%
  rename(TSTOP = TIME_TO_EVENT) %>%
  select(ID, TREAT, TSTART, TSTOP, EVENT)    # Keep relevant columns for events

# Reshape data for censoring using survSplit
combined_data_cens <- survSplit(
  Surv(TSTART, TIME_TO_CENSOR, CENSOR) ~ ., 
  data = combined_data, 
  cut = t_censor, 
  id = "id"
) %>%
  rename(TSTOP = TIME_TO_CENSOR) %>%
  select(ID, TREAT, TSTART, TSTOP, CENSOR)  

# Extract covariates for the analysis
combined_data_covariate <- combined_data %>%
  select(ID, AGE, GENDER, SOCIO_ECO_2, SOCIO_ECO_3, SOCIO_ECO_4, SOCIO_ECO_5, BECK)

# Combine event and censoring datasets
combined_data <- full_join(
  combined_data_events,
  combined_data_cens,
  by = c("ID", "TREAT", "TSTART", "TSTOP")  # Merge by ID, treatment, start, and stop times
) %>%
  mutate(
    EVENT = ifelse(is.na(EVENT), 0, EVENT),   # Replace missing EVENT values with 0
    CENSOR = ifelse(is.na(CENSOR), 0, CENSOR) # Replace missing CENSOR values with 0
  ) %>%
  group_by(ID) %>%
  mutate(
    censor_cumsum = cumsum(cumsum(CENSOR)),   # Count cumulative censored events per ID
    event_cumsum = cumsum(cumsum(EVENT))      # Count cumulative relapse events per ID
  ) %>%
  filter(censor_cumsum <= 1 & event_cumsum <= 1) %>%  # Retain only one censor or event per ID
  select(-censor_cumsum, -event_cumsum) %>%
  inner_join(combined_data_covariate, by = "ID")

combined_data$TREAT<-factor(combined_data$TREAT,levels=c("Sertralex","Duloxyn"))

# Print first individual (ID = 3) and its clone (ID = 1003)
combined_data %>%
  filter(ID %in% c(3,1003)) %>%
  arrange(ID) %>%
  kbl() %>%
  kable_styling(font_size = 12) %>%
  scroll_box(width = "100%", height = "200px")
```

### Weighting

Finally, the probability of remaining uncensored is estimated separately using a Cox model.

```{r weighting, message=FALSE}
# Step 3 weighting

#Filter data for the "Duloxyn" treatment group and estimate censoring probabilities
combined_data_duloxyn <- combined_data %>% filter(TREAT == "Duloxyn")

# Fit a Cox proportional hazards model to estimate the hazard for censoring
ms_cens_duloxyn <- coxph(Surv(TSTART, TSTOP, CENSOR) ~ AGE + GENDER + SOCIO_ECO_2 + SOCIO_ECO_3 + SOCIO_ECO_4 + SOCIO_ECO_5 + BECK, 
                         ties = "efron", data = combined_data_duloxyn)

# Extract the baseline hazard from the Cox model
baseline_hazard <- basehaz(ms_cens_duloxyn, centered = FALSE)

# Calculate the linear predictor (lp) for the "Duloxyn" group
combined_data_duloxyn$lp <- predict(ms_cens_duloxyn, newdata = combined_data_duloxyn, type = "lp")

# Merge the baseline hazard with the dataset based on time
combined_data_duloxyn <- merge(combined_data_duloxyn, baseline_hazard, 
                               by.x = "TSTOP", by.y = "time", all.x = TRUE)

# Calculate the uncensored probability
combined_data_duloxyn$p_uncens <- exp(-combined_data_duloxyn$hazard * exp(combined_data_duloxyn$lp))

# Calculate the weights as the inverse of the uncensored probability
combined_data_duloxyn$weights <- 1 / combined_data_duloxyn$p_uncens

# Repeat the same process for the "Sertralex" treatment group
# Filter data for the "Sertralex" treatment group

combined_data_sertralex <- combined_data %>% filter(TREAT == "Sertralex")

# Fit a Cox proportional hazards model for censoring in the "Sertralex" group
ms_cens_sertralex <- coxph(Surv(TSTART, TSTOP, CENSOR) ~ AGE + GENDER + SOCIO_ECO_2 + SOCIO_ECO_3 + SOCIO_ECO_4 + SOCIO_ECO_5 + BECK, 
                           ties = "efron", data = combined_data_sertralex)

# Use the baseline hazard from the "Duloxyn" model (assumed shared here)
baseline_hazard <- basehaz(ms_cens_duloxyn, centered = FALSE)

# Calculate the linear predictor (lp) for the "Sertralex" group
combined_data_sertralex$lp <- predict(ms_cens_sertralex, newdata = combined_data_sertralex, type = "lp")

# Merge the baseline hazard with the dataset based on time
combined_data_sertralex <- merge(combined_data_sertralex, baseline_hazard, 
                                 by.x = "TSTOP", by.y = "time", all.x = TRUE)

# Calculate the uncensored probability
combined_data_sertralex$p_uncens <- exp(-combined_data_sertralex$hazard * exp(combined_data_sertralex$lp))

# Calculate the weights as the inverse of the uncensored probability
combined_data_sertralex$weights <- 1 / combined_data_sertralex$p_uncens

# Combine the datasets for "Duloxyn" and "Sertralex" treatments with the calculated weights
combined_data <- rbind(combined_data_duloxyn, combined_data_sertralex)

# Display the first individual (ID = 3) and its clone (ID = 1003) with improved visualization
combined_data %>%
  filter(ID %in% c(3, 1003)) %>%
  arrange(ID) %>%
  kbl() %>%
  kable_styling(font_size = 12) %>%
  scroll_box(width = "100%", height = "200px")
```

## Outcome analysis

First, the KM model is implemented.

```{r ccw_km, warning=FALSE}
#Implement CCW  model
ccw<- survfit(Surv(TSTART, TSTOP, EVENT) ~ TREAT, data = combined_data, weights = weights)

# Calculate median survival time
ccw_treatment_0 <- format_median_time(ccw, 1)
ccw_treatment_1 <- format_median_time(ccw, 2)

ggsurvplot(ccw, 
           data = combined_data,
           legend.title = "",
           legend.labs = c("CCW: Sertralex",
                           "CCW: Duloxyn"),
           palette = c("#5A8DEE","#F28C2C"),
           conf.int = TRUE,
           xlab = "Time (in days)",
           ggtheme = theme_bw())
```

In the CCW approach, the estimated median survival time in the Sertralex population is estimated at `r ccw_treatment_0` and `r ccw_treatment_1` in the Duloxyn population.

Next, the Cox model is implemented where weights are considered into the analysis.

```{r ccw_cox}
# Fit the Cox models
combined_data$TREAT <- relevel(combined_data$TREAT,ref="Sertralex")
CCW_cox <- coxph(Surv(TSTART, TSTOP, EVENT) ~ TREAT, data = combined_data, weights = weights)

# Create regression tables
CCW_cox %>% 
  tbl_regression(exponentiate = TRUE,
                 label = list(TREAT ~ "Treatment (CCW approach)"))

ccw_hr_ci <- extract_hr_ci(CCW_cox)
```

After adjusting for immortal time bias, the estimated treatment effect is represented by `r ccw_hr_ci`, indicating that the association becomes significant. This finding aligns with the simulation process, which modeled a average difference in treatment initiation of 5 days (15 days for Duloxyn and 10 days for Sertralex), thereby offering a survival advantage for Duloxyn users.